Vagueness and imprecise imitation in signaling games

### 2016-03-03 paper/version_05/paper.tex
replies to reviewers
final tweaks responses and paper

Formally, a sim-max game consists of a set of states $\States$, a set of messages $\Messgs$
typically with much fewer messages than states, a probability distribution
$\Pr \in \Delta(\States)$ such that $\Pr(\state)$ gives the probability that state $\state$
occurs, a similarity metric\footnote{A metric is a function of distance between any two points
  in a given space. It should satisfy certain axioms that ensure behavior that one would
  intuitively expect from the words ``similarity'' and ``distance'' alone, but these details do
  not matter for the purposes of this paper. The assumption that state similarity forms a
  metric is conceptually loaded, but we follow the literature here and conceive of it as a
  first and pragmatic simplification, possibly to be dispensed with later.}
  % : similarity should always be non-negative, zero only for the distance
  % between an element and itself, symmetric, and satisfy the triangle inequality that
  % $\similarity(x,z) \le \similarity(x,y) + \similarity(y,z)$
on states $\similarity \mycolon \States \times \States \rightarrow \mathds{R}$ such that
$\similarity(\mystate{1}, \mystate{2})$ is the (perceptual) similarity between $\mystate{1}$
and $\mystate{2}$, and a utility function
$\Utils \mycolon \States \times \States \rightarrow \mathds{R}$ such that
$\Utils(\mystate{1}, \mystate{2})$ is the payoff for sender and receiver for a play with actual
state $\mystate{1}$ and receiver interpretation $\mystate{2}$. We identify the receiver's acts
with the states of the world, so that the game is one of guessing the actual state, so to
speak. The utility function should be a monotonically increasing function of
similarity.\footnote{Section~\ref{sec:setting-stage} motivates a particular choice of utility
  function that we will explore in more detail. But since the general discussion also applies
  and makes reference to other choices, we prefer to remain general for the time being.}
Non-probabilistic sender behavior can be represented by a pure strategy
$\Spure \in \Messgs^\States$, which deterministically defines which message would be used for
each state. Similarly, a pure receiver strategy is a function $\Rpure \in \States^\Messgs$.

@@

Alternative approaches are taken by, for example, \citet{FrankeJager2010:Vagueness-Signa} and
\citet{OConnor2013:The-Evolution-o}, who show different ways how the above desiderata can be
met by evolving strategies in sim-max games. To illustrate what vague signaling would look
like, let us briefly consider a sim-max game with six equiprobable states and two messages and
what its equilibria would be like \citep[see][for a more thorough discussion of equilibria of
sim-max games]{Jager2007:The-Evolution-o,OConnor2013:The-Evolution-o}. We assume that utilities
are linearly decreasing with decreasing similarity. Figure~\ref{fig:EquilibriaVagueness} shows
three pairs of sender and receiver strategies. States are arranged according to their
similarity: the closer they are to each other, the more similar they are. The pair in
Figure~\ref{fig:ineffNonEq} is not an equilibrium, because the sender's non-convex use of
signals is suboptimal given the receiver's behavior. Namely, if \mymessg{1} is interpreted as
\mystate{2} and \mymessg{2} as \mystate{4}, then the sender would get a higher payoff from
sending \mymessg{1} in \mystate{3} than from sending \mymessg{2}, because (by assumption)
\mystate{3} is more similar to \mystate{2} than \mystate{5} is. In contrast,
Figure~\ref{fig:efficientEq} shows a maximally efficient equilibrium. This is a partial pooling
equilibrium in the sense that the speaker uses the same message for several states. Partial
pooling equilibria can be less inefficient than other non-pooling equilibria, but that requires
enough messages. The strategy pair of Figure~\ref{fig:efficientEq} is maximally efficient for
the two-message case. So, while partial pooling may entail inefficiency in some sense, and
while partial pooling can hamper the evolution of maximally efficient signal use
\citep[e.g.][]{Hutteger:2007_Evol_Indicatives_Imperatives,Pawlowitsch2008:Why-Evolution-d,HutteggerSkyrms2010:Evolutionary-Dy},
this is orthogonal to our concerns about vagueness. A regular and natural yet vague signal use
would look like the pair in Figure~\ref{fig:vagueUse}. This is not an equilibrium, but it gets
close, so to speak. It shows smooth transitions across similar states at the boundaries of
categories and across acts around the most prototypical instances (as indicated by decreasing
thickness of arrows in Figure~\ref{fig:vagueUse}).

@@

The emphasis of previous models that studied the effects of transmission infidelity on the
evolution of language has been on inductive biases and the systematicity, compression and
regularization that they can introduce. Here, we would like to show that transmission noise of
a different kind can lead to regularization as well and also give rise to vague meaning. While
perhaps \emph{prima facie} unexpected, we find that shared perceptual biases that perturb the
transmission of successful signaling behavior can regularize, facilitate and accelerate the
evolution of meaning conventions, albeit at the cost of vagueness. To show this, we formalize
the expected change in the behavior of a population of agents that try to imitate other agents'
signaling behavior. We assume, however, that both realization of behavior, as well as
observability of others' behavior, is systematically perturbed by noise. The resulting
population-level dynamic generalizes the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St} in its interpretation as a cultural evolutionary
dynamic based on imitation
\citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and}. The inclusion of
confusability of stimuli does not undermine the possibility of evolving communicative signaling
behavior. Instead, it leads to the evolution of vague meanings. It moreover accelerates the
emergence of communicative signaling because it unifies and regularizes evolutionary outcomes,
making it appear as if agents were applying inductive biases or generalizing over partial
observations, when in truth this is the sole effect of confusion of similar stimuli.

@@

Appendix~\ref{sec:imit-succ-with} shows how imprecise imitation of this sort leads to mean changes in population
frequencies of choices that can be covered by the following discrete-time formulation:
\begin{align}
  \label{eq:nRD}
  \Sstrat'(\messg \probbar \state) & \propto P_o(\messg \probbar \state) \myts \EU(\messg,
  \state, \Rstrat)\,, & \Rstrat'(\state \probbar \messg) & \propto P_o(\state \probbar \messg)
  \myts \EU(\state, \messg, \Sstrat)\,.
\end{align}
\noindent This looks very much like the discrete-time formulation of the standard replicator
dynamic in (\ref{eq:RDDiscrete}), but there are, of course, the aforementioned
differences. Firstly, expected utility here takes stochastic confusability of states into
account. Secondly, where the standard replicator dynamic had probabilities
$\Sstrat(\messg \probbar \state)$ and $\Rstrat(\state \probbar \messg)$, we now have
$P_o(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ respectively. If there is no
observation or realization noise, $P_o(\messg \probbar \state)$ reduces to
$\sigma(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ reduces to
$\Rstrat(\state \probbar \messg)$. The noisy replicator dynamic in (\ref{eq:nRD})
conservatively extends the classic case in (\ref{eq:RDDiscrete}).

>>>>

Formally, a sim-max game consists of a set of states $\States$, a set of messages $\Messgs$
typically with much fewer messages than states, a probability distribution
$\Pr \in \Delta(\States)$ such that $\Pr(\state)$ gives the probability that state $\state$
occurs, a similarity metric\footnote{A metric is a function of distance between any two points
  in a given space. It should satisfy certain axioms that ensure behavior that one would
  intuitively expect from the words ``similarity'' and ``distance'' alone, but these details do
  not matter for the purposes of this paper. The assumption that state similarity forms a
  metric is conceptually loaded, but we follow the literature here and conceive of it as a
  first and pragmatic simplification, possibly to be dispensed with later.}
  % : similarity should always be non-negative, zero only for the distance
  % between an element and itself, symmetric, and satisfy the triangle inequality that
  % $\similarity(x,z) \le \similarity(x,y) + \similarity(y,z)$
on states $\similarity \mycolon \States \times \States \rightarrow \mathds{R}$ such that
$\similarity(\mystate{1}, \mystate{2})$ is the (physical) similarity between $\mystate{1}$ and
$\mystate{2}$, and a utility function
$\Utils \mycolon \States \times \States \rightarrow \mathds{R}$ such that
$\Utils(\mystate{1}, \mystate{2})$ is the payoff for sender and receiver for a play with actual
state $\mystate{1}$ and receiver interpretation $\mystate{2}$. We identify the receiver's acts
with the states of the world, so that the game is one of guessing the actual state, so to
speak. For the modeling purposes of this paper, we make the simplifying assumption that
$\States$ is a set of points in Euclidean space, whose closeness to each other tracks physical
similarity. Perceived similarity, where it is necessary, would be a monotonic function of
physical similarity. Likewise, the utility function should be a monotonically increasing
function of physical similarity.\footnote{Section~\ref{sec:setting-stage} motivates particular
  choices of similarity and utility functions that we will explore in more detail.}
Non-probabilistic sender behavior can be represented by a pure strategy
$\Spure \in \Messgs^\States$, which deterministically defines which message would be used for
each state. Similarly, a pure receiver strategy is a function $\Rpure \in \States^\Messgs$.

@@

Alternative approaches are taken by, for example, \citet{FrankeJager2010:Vagueness-Signa} and
\citet{OConnor2013:The-Evolution-o}, who show different ways how the above desiderata can be
met by evolving strategies in sim-max games. To illustrate what vague signaling would look
like, let us briefly consider a sim-max game with six equiprobable states and two messages and
what its equilibria would be like \citep[see][for a more thorough discussion of equilibria of
sim-max games]{Jager2007:The-Evolution-o,OConnor2013:The-Evolution-o}. We assume that utilities
are linearly decreasing with decreasing similarity. Figure~\ref{fig:EquilibriaVagueness} shows
three pairs of sender and receiver strategies. States are arranged according to their
similarity: the closer they are to each other, the more similar they are. The pair in
Figure~\ref{fig:ineffNonEq} is not an equilibrium, because the sender's non-convex use of
signals is suboptimal given the receiver's behavior. Namely, if \mymessg{1} is interpreted as
\mystate{2} and \mymessg{2} as \mystate{4}, then the sender would get a higher payoff from
sending \mymessg{1} in \mystate{3} than from sending \mymessg{2}, because (by assumption)
\mystate{3} is more similar to \mystate{2} than \mystate{5} is. In contrast,
Figure~\ref{fig:efficientEq} shows a maximally efficient equilibrium. This is a partial pooling
equilibrium in the sense that the speaker uses the same message for several states. Partial
pooling equilibria can be less inefficient than other non-pooling equilibria, if there are
enough messages. The strategy pair of Figure~\ref{fig:efficientEq} is maximally efficient for
the two-message case. So, while partial pooling may entail inefficiency in some sense, and
while partial pooling can hamper the evolution of maximally efficient signal use
\citep[e.g.][]{Hutteger:2007_Evol_Indicatives_Imperatives,Pawlowitsch2008:Why-Evolution-d,HutteggerSkyrms2010:Evolutionary-Dy},
this is orthogonal to our concerns about vagueness. A regular and natural yet vague signal use
would look like the pair in Figure~\ref{fig:vagueUse}. This is not an equilibrium, but it gets
close, so to speak. It shows smooth transitions across similar states at the boundaries of
categories and across acts around the most prototypical instances (as indicated by decreasing
thickness of arrows in Figure~\ref{fig:vagueUse}).

@@

The emphasis of previous models that studied the effects of transmission infidelity on the
evolution of language has been on inductive biases and the systematicity, compression and
regularization that they can introduce. Here, we would like to show that transmission noise of
a different kind can lead to regularization as well and also give rise to vague meaning. We
find that shared perceptual biases that perturb the transmission of successful signaling
behavior can regularize, facilitate and accelerate the evolution of meaning conventions, albeit
at the cost of vagueness. Concretely, we formalize the expected change in the behavior of a
population of agents that try to imitate other agents' signaling behavior. We assume, however,
that both realization of behavior, as well as observability of others' behavior, is
systematically perturbed by noise. The resulting population-level dynamic generalizes the
replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} in its interpretation as a cultural
evolutionary dynamic based on imitation
\citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and}. The inclusion of
confusability of stimuli does not undermine the possibility of evolving communicative signaling
behavior. Instead, it leads to the evolution of vague meanings. It moreover accelerates the
emergence of communicative signaling because it unifies and regularizes evolutionary outcomes,
making it appear as if agents were applying inductive biases or generalizing over partial
observations, when in truth this is the sole effect of confusion of similar stimuli.

@@

Appendix~\ref{sec:imit-succ-with} shows how imprecise imitation of this sort leads to mean changes in population
frequencies of choices that can be covered by the following discrete-time formulation:
\begin{align}
  \label{eq:nRD}
  \Sstrat'(\messg \probbar \state) & \propto P_o(\messg \probbar \state) \myts \EU(\messg,
  \state, \Rstrat)\,, & \Rstrat'(\state \probbar \messg) & \propto P_o(\state \probbar \messg)
  \myts \EU(\state, \messg, \Sstrat)\,.
\end{align}
\noindent This looks very much like the discrete-time formulation of the standard replicator
dynamic in (\ref{eq:RDDiscrete}), but there are, of course, the aforementioned
differences. Firstly, expected utility here takes stochastic confusability of states into
account. Secondly, where the standard replicator dynamic had probabilities
$\Sstrat(\messg \probbar \state)$ and $\Rstrat(\state \probbar \messg)$, we now have
$P_o(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ respectively. If there is no
observation or realization noise, $P_o(\messg \probbar \state)$ reduces to
$\sigma(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ reduces to
$\Rstrat(\state \probbar \messg)$. The imprecise imitation dynamic in (\ref{eq:nRD})
conservatively extends the classic case in (\ref{eq:RDDiscrete}).
